---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First load the libraries needed in this assessment

```{r,echo=TRUE}
library(dplyr)
library(lubridate)
library(lattice)
```

Read in the data set and remove the rows with steps=NA

```{r,echo=TRUE}
activity <- read.csv("./activity.csv")
activity2 <- filter(activity,is.na(steps)==FALSE)
```



## What is mean total number of steps taken per day?

Calculate the total number of steps each day and produce a histogram
of the results

```{r first_hist,echo=TRUE}
activity2 <- group_by(activity2,date)
act_per_day <- summarize(activity2,steps_per_day = sum(steps))
hist(act_per_day$steps_per_day,main="Histogram of Total Steps per Day",xlab = "Total Steps per Day")
```

Calculate and report the mean and median of the total number of steps taken per day

```{r,echo=TRUE}
mn <- mean(act_per_day$steps_per_day)
med <- median(act_per_day$steps_per_day)
```

The mean steps per day is `r mn`.
The median steps per day is `r med`.

## What is the average daily activity pattern?
Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r first_time_series,echo=TRUE}
activity3 <- filter(activity,is.na(steps)==FALSE)
activity3 <- group_by(activity3,interval)
act_by_time <- summarize(activity3,steps_by_time = mean(steps))
plot(act_by_time,type="l")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r,echo=TRUE}
max_start <-act_by_time$interval[act_by_time$steps_by_time==max(act_by_time$steps_by_time)]
max_fin <- max_start + 5
```

The period with the maximum number of steps is between `r max_start` and `r max_fin` minutes


## Imputing missing values

Calculate and report the total number of missing values in the dataset 

```{r,echo=TRUE}
activity_na <- filter(activity,is.na(steps)==TRUE)
activity_good <- filter(activity,is.na(steps)==FALSE)
missing <- nrow(activity_na)
```

There are `r missing` rows with missing values in the dataset


Fill in the missing values by setting them equal to the mean for that 5 minute period over all days

```{r,echo=TRUE}
activity_na <- left_join(activity_na,act_by_time,by="interval")
activity_na$steps <- round(activity_na$steps_by_time)
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.


```{r,echo=TRUE}
activity_na <- select(activity_na,-steps_by_time)
activity_na$steps <- as.integer(activity_na$steps)
activity_na$interval <- as.integer(activity_na$interval)
activity_all <- rbind(activity_na,activity_good)
activity_all <- arrange(activity_all,date,interval)
```

Make a histogram of the total number of steps taken each day now that the missing values have been filled in.

```{r second_hist,echo=TRUE}
activity_all <- group_by(activity_all,date)
act_per_day2 <- summarize(activity_all,steps_per_day = sum(steps))
hist(act_per_day2$steps_per_day,main="Histogram of Total Steps per Day (Missing Values Replaced)",xlab = "Total Steps per Day")
```

Calculate and report the mean and median total number of steps taken per day.

```{r,echo=TRUE}
mn2 <- mean(act_per_day2$steps_per_day)
med2 <- median(act_per_day2$steps_per_day)
```

The mean steps per day (now that missing values have been filled in) is `r mn2`.
The median steps per day (now that missing values have been filled in) is `r med2`.


Calculate the impact of imputing missing data on the estimates of the total daily number of steps by comparing with the previous means and medians we calculated.

```{r,echo=TRUE}
diffmean <- mn2-mn
diffmed <- med2 - med
```

The difference in the means by filling in missing values is `r diffmean`. 
The difference in the medians by filling in missing values is `r diffmed`


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.


```{r,echo=TRUE}
activity_all$date <- ymd(activity_all$date)
activity_all$day <- weekdays(activity_all$date)
activity_all$weekend[activity_all$day == "Saturday" | activity_all$day == "Sunday"] = "weekend"
activity_all$weekend[is.na(activity_all$weekend) == TRUE] <- "weekday"

```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r second_time_series,echo=TRUE}
activity_all <- group_by(activity_all,interval,weekend)
activity_weekend <- summarize(activity_all,steps = mean(steps))
xyplot(steps~interval|weekend,activity_weekend,type="l")

```

